---
title: "Herring recruitment and SSB analyses for paper"
date: "03-28-2016"
output:
  pdf_document:
    fig_caption: yes
  html_document: default
  word_document: default
---

This is supplementary information for the Ward et al. paper 

# Data Processing  
```{r,echo=FALSE,message=FALSE}
# Try readxl package
if("readxl" %in% rownames(installed.packages()) == FALSE) {
  devtools::install_github("hadley/readxl")}
require(readxl)
library(MARSS)
library(knitr)
```

### Read in the data. We'll primarily use data from brood years 1981 - 2011, because those are the span of years from the ASA model with R/S and covariates known (hatchery releases before 1980 incomplete).
```{r}
pwsher = read_excel("../../data/herring/PWS_herring_final.xlsx")
```

Plot the response, log(Recruits/Spawners).
```{r,echo=TRUE,fig.cap='log(Recruits / Spawner) over time, 1981-2011',fig.pos="placeHere"}
subset = which(pwsher$BroodYear%in%seq(1981,2011))
Y = log(as.numeric(pwsher$RecPerSpawn[subset])) # log(R/S)

par(mfrow = c(2,2),mgp=c(2,1,0),mai=c(0.8,0.6,0.2,0.05))
plot(1981:2011,Y, xlab="Year",ylab="log(Age.3 Recruits/Spawner)",main="PWS",type="b")
plot(1981:2011, pwsher$Rec3Obs[subset], xlab="Year",ylab="Recruits",main="PWS",type="b")
plot(1981:2011, pwsher$BroodYearSB[subset], xlab="Year",ylab="Spawners",main="PWS",type="b")
```

Plot the data, as log(Recruits/Spawners) versus Spawners over the period we're using, 1981-2011. This is the same formulation as the Ricker model assumes (below).
```{r,fig.cap='Raw data plot of log(R/S) on Spawners, 1981-2011',fig.pos="placeHere"}
subset = which(pwsher$BroodYear%in%seq(1981,2011))
Y = log(as.numeric(pwsher$RecPerSpawn[subset])) # log(R/S)
X = as.numeric(pwsher$BroodYearSB[subset]) # number of spawners

par(mfrow = c(2,2),mgp=c(2,1,0),mai=c(0.8,0.6,0.2,0.05))
plot(X,Y, xlab="Spawners",ylab="log(Recruits/Spawner)",main="",type="b")
# fit linear model
lines(X, predict.lm(lm(Y~X),newdata=data.frame(X)), col="blue",lwd=3)
mod = lm(Y~X)
plot(1981:2011,mod$residuals, xlab="Year",ylab="Residuals",main="",type="b")
```

# Modeling recruiutment  

We'll conduct this analysis using the Ricker stock-recruit model, which is equivalent to a linear regression model,  
$log(R/S)_{t} = {a}_{i} + {b}_{i}*{S}_{t} + {c}_{i}*{X}_{t}  + {e}_{i}$  
  
where ${a}_{i}$ represents the population-specific intercept, ${b}_{i}$ is a density-dependent parameter (generally negative), ${c}_{i}$ is an optional coefficient(s) incorporating a time-varying covariate ${X}_{t}$, and ${e}_{i}$ is an error term. Simple models for the error are IID white noise, which we'll adopt here.

## Constructing the basic (null) model with no covariates. 
We'll start with just using data 1981-2011, and spawners as a predictor of recruitment. Recruits / spawner is not modeled as an autoregressive state-space process, but all uncertainty is assumed to have arisen from measurement and observation error. Hypotheses for mechanistic relationships are discussed and evaluated below. 

```{r,warning=FALSE,silent=TRUE}
subset = which(pwsher$BroodYear%in%seq(1981,2011))

Y = log(as.numeric(pwsher$RecPerSpawn[subset])) # log(R/S)
X = as.numeric(pwsher$RecPerSpawn[subset]) # number of spawners
nT = length(Y)

# fit in initial Ricker S-R state space model 
cMat = matrix(NA, nrow=1, ncol = nT)
cMat[1,] = X

models = list()
```

```{r,warning=FALSE,silent=TRUE,echo=FALSE,results="hide"}
models[[1]] = MARSS(Y, model = list("Q"="zero"))
models[[2]] = MARSS(Y, model = list("d"=cMat,"Q"="zero"))
m = matrix(NA,2,3)
colnames(m) = c("Model","AICc","Coef")
m[1,1]="Null model"
m[1,2]=round(models[[1]]$AICc,3)
m[2,1]="Density dependence"
m[2,2]=round(models[[2]]$AICc,3)
m[2,3]=round(models[[2]]$par$A,8)
```

```{r,echo=FALSE}
kable(m)
```

# Hypothesis 2: EVOS had an impact on herring productivity

The EVOS spill occurred in 1989. Herring typically migrate to the ocean 2 years after spawning, so the immediate impacts of the spill may have impacted recruitment from brood years 1987, 1988, and 1989. 
  
We'll include the impacts of the EVOS spill. We'll do this 3 ways: creating a pulse impact, a press impact, and a press impact followed by a recovery back to the original state. The form of the recovery was assumed to be linear over a 20 - year period.  

[Note: a negative coefficient on the press or pulse corresponds to a negative impact; because of how we coded the dummy covariate, a negative coefficient on the pulse-recovery change translates into a positive perturbation]  

```{r,fig.cap="Illustration of covariates representing EVOS impacts",fig.pos="placeHere"}
par(mfrow = c(2,2),mgp=c(2,1,0))
plot(pwsher$BroodYear, 1-pwsher$EVOS.pulse.lag0, xlab = "", ylab = "Impact",
     main = "Pulse",col="blue",lwd=3,type="l")
plot(pwsher$BroodYear, 1-pwsher$EVOS.press.lag0, xlab = "", ylab = "Impact",
     main = "Press",col="blue",lwd=3,type="l")
plot(pwsher$BroodYear, pwsher$EVOS.pulseRecovery.lag0, xlab = "", ylab = "Impact",
     main = "Pulse/Recovery",col="blue",lwd=3,type="l")
```

```{r,warning=FALSE}
library(MARSS)
covar.names = c("EVOS.pulse.lag0", "EVOS.press.lag0", "EVOS.pulseRecovery.lag0", 
    "EVOS.pulse.lag1", "EVOS.press.lag1", "EVOS.pulseRecovery.lag1",  
"EVOS.pulse.lag2", "EVOS.press.lag2", "EVOS.pulseRecovery.lag2")
# fit in initial Ricker S-R state space model 

cMat = matrix(NA, nrow=2, ncol = nT)
cMat[1,] = X

evos.models = list()
```

```{r,echo=FALSE,warning=FALSE,silent=TRUE,results="hide"}
m = matrix(NA,length(covar.names),3)
colnames(m) = c("Model","AICc","Coef")
m[,1] = covar.names

for(i in 1:length(covar.names)) {
# include this covariate as predictor
cMat[2,] = pwsher[subset,covar.names[i]]
# set Q = 0, observation error only
evos.models[[i]] = MARSS(Y, model = list("d"=cMat,"Q"="zero"))
m[i,2] = round(evos.models[[i]]$AICc,3)
m[i,3] = round(evos.models[[i]]$par$A[2],8)
}
```

```{r,echo=FALSE}
kable(m)
```

These results show that most of the EVOS models do worse than the null model, maybe with the exception of the lag.0 model. 

```{r,echo=FALSE}
#kable(m)
```
\newpage  
  
# Hypothesis 4: Herring productivity in PWS has been affected by predation and competition from juvenile pink salmon.  
Age-1 herring in 1969 may be affected by pink salmon released in 1969 (BY 1968), competing in the later summer or fall months.

```{r,warning=FALSE}
library(MARSS)
covar.names = c("juv.hatchRelPink.lag2","juv.hatchRelChum.lag2",
"juv.wildPinkRun.lag1","juv.wildChumRun.lag2")
# fit in initial Ricker S-R state space model 

cMat = matrix(NA, nrow=2, ncol = nT)
cMat[1,] = X

juvComp.models = list()
```

```{r,echo=FALSE,warning=FALSE,silent=TRUE,results="hide"}
m = matrix(NA,length(covar.names),3)
colnames(m) = c("Model","AICc","Coef")
m[,1] = covar.names

for(i in 1:length(covar.names)) {
# include this covariate as predictor
cMat[2,] = pwsher[subset,covar.names[i]]
# set Q = 0, observation error only
juvComp.models[[i]] = MARSS(Y, model = list("d"=cMat,"Q"="zero"))
m[i,2] = round(juvComp.models[[i]]$AICc,3)
m[i,3] = round(juvComp.models[[i]]$par$A[2],8)
}

```
  
```{r,echo=FALSE}
kable(m)
```
  
These results show that there isn't much support for including pink or chum competion with herring as a predictor for the decline (AICc worse than null model).
  
# Hypothesis 5: Herring productivity in PWS has been affected by predation and competition from adult pink (and chum) salmon

```{r,warning=FALSE}
library(MARSS)
pwsher$ad.totalPinkRun.lag1 = pwsher$ad.wildPinkRun.lag1 + pwsher$ad.hatchPinkRun.lag1
covar.names = c("ad.hatchPinkRun.lag1","ad.hatchRelChum.lag2",
"ad.wildPinkRun.lag1","ad.wildChumRun.lag1",
"ad.totalPinkRun.lag1")
# fit in initial Ricker S-R state space model 

cMat = matrix(NA, nrow=2, ncol = nT)
cMat[1,] = X

adComp.models = list()
```

```{r,echo=FALSE,warning=FALSE,silent=TRUE,results="hide"}
m = matrix(NA,length(covar.names),3)
colnames(m) = c("Model","AICc","Coef")
m[,1] = covar.names
 
for(i in 1:length(covar.names)) {
# include this covariate as predictor
cMat[2,] = pwsher[subset,covar.names[i]]
# set Q = 0, observation error only
adComp.models[[i]] = MARSS(Y, model = list("d"=cMat,"Q"="zero"))
m[i,2] = round(adComp.models[[i]]$AICc,3)
m[i,3] = round(adComp.models[[i]]$par$A[2],8)
}
```

```{r,echo=FALSE}
kable(m)
```

These results show that adult runs of chum or pink salmon may have a negative impact on PWS herring recruitment.

# Hypothesis 3: Herring productivity in PWS has been shaped by changing ocean

```{r,warning=FALSE}
library(MARSS)
covar.names = c("humpbacks","Upwelling.summerBefore",
"Upwelling.summerAfter","discharge.lag0","discharge.lag1",
"win.sst.lag1","win.sst.lag0")
# fit in initial Ricker S-R state space model 

cMat = matrix(NA, nrow=2, ncol = nT)
cMat[1,] = X

enviro.models = list()
```

```{r,echo=FALSE,warning=FALSE,silent=TRUE,results="hide"}
m = matrix(NA,length(covar.names),3)
colnames(m) = c("Model","AICc","Coef")
m[,1] = covar.names

for(i in 1:length(covar.names)) {
# include this covariate as predictor
cMat[2,] = pwsher[subset,covar.names[i]]
# set Q = 0, observation error only
enviro.models[[i]] = MARSS(Y, model = list("d"=cMat,"Q"="zero"),control=list("maxit"=1000))
m[i,2] = round(enviro.models[[i]]$AICc,3)
m[i,3] = round(enviro.models[[i]]$par$A[2],8)
}

```

```{r,echo=FALSE}
kable(m)
```

These results show that most of the environmental indices (or humpbacks) doesn't appear to have a strong impact on PWS herring recruitment. The exception is discharge, which has a huge improvement as a predictor (- 8 log likelihood units)

Using this best model, let's look at including some of the EVOS or pink/chum covariates with discharge.

```{r,warning=FALSE}
library(MARSS)
covar.names = c("EVOS.press.lag2","EVOS.pulseRecovery.lag2",
"ad.hatchRelPink.lag2","ad.hatchRelChum.lag2")

cMat = matrix(NA, nrow=3, ncol = nT)
cMat[1,] = X
cMat[2,] = pwsher[subset,"discharge.lag0"]

combo.models = list()
```

```{r,echo=FALSE,warning=FALSE,silent=TRUE,results="hide"}
m = matrix(NA,length(covar.names),3)
colnames(m) = c("Model","AICc","Coef")
m[,1] = paste("Discharge+",covar.names)

for(i in 1:length(covar.names)) {
# include this covariate as predictor
cMat[3,] = pwsher[subset,covar.names[i]]
# set Q = 0, observation error only
combo.models[[i]] = MARSS(Y, model = list("d"=cMat,"Q"="zero"),control=list("maxit"=1000))
m[i,2] = round(combo.models[[i]]$AICc,3)
m[i,3] = round(combo.models[[i]]$par$A[2],3)
}
```

```{r,echo=FALSE}
kable(m)
```

What this shows is that once all of these predictors to the model with discharge, they do roughly equally well, so the model with discharge alone is still best. 

The R^2 from this model is about ~ 0.55, and we can look at the plot of observed and predicted values.

```{r,warning=FALSE}

cMat = matrix(NA, nrow=2, ncol = nT)
cMat[1,] = X
cMat[2,] = pwsher[subset,"discharge.lag0"]
mod = lm(Y~cMat[1,]+cMat[2,])
```

```{r,fig.cap="Observed (red points) and predicted (black line) recruits per spawner, 1981-2011",fig.pos="placeHere"}

pdf("Figure 6 herring.pdf")
expr = expression(paste("Total discharge ", m^3, " ", s^-1, sep=""))
par(mfrow=c(2,1),mgp=c(2,1,0),mai=c(0.7,0.7,0.3,0.1))
plot(pwsher$BroodYear[which(pwsher$BroodYear%in%seq(1981,2011))],pwsher$discharge.lag0[which(pwsher$BroodYear%in%seq(1981,2011))], xlab="", ylab=expr,
     type="b", lwd=3, xlim=c(1981,2011))
legend('topleft',"(a)", bty='n')
plot(1981:2011, lm(Y~cMat[1,]+cMat[2,])$fitted.values, xlab="", ylim=c(-16,1),type="l",lwd=3,ylab="log (age 3 recruits / SSB)")
legend('topleft',"(b)", bty='n')
points(1981:2011, Y, col="grey30",lwd=2,cex=1.2)
dev.off()
```
